Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-20 paper/paper.Rmd
adds creating maps section

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: maximilian.hesselbarth@uni-goettingen.de
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: | 
  \usepackage{caption}
  \usepackage{ltablex}
  \usepackage{lscape}
  \usepackage{multirow}
output: rticles::springer_article
editor_options: 
  chunk_output_type: console
---

@@

## Spatial data representations

@@

<!-- role of making maps - from checking data, through the data exploration, to presenting the results  -->

<!-- maps in the computer age -->
<!-- static maps and interactive maps -->

<!-- build-in plotting functions -->
<!-- tmap -->
<!-- leaflet/mapview -->
<!-- mention ggplot2/ggspatial as well -->

>>>>

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: maximilian.hesselbarth@uni-goettingen.de
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: | 
  \usepackage{caption}
  \usepackage{ltablex}
  \usepackage{lscape}
  \usepackage{multirow}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
editor_options: 
  chunk_output_type: console
---

@@

## Spatial data representations {#spatrep}

@@

Creating maps is essential when working with spatial data.
It has a role in checking the quality of data<!--(both coordinates and values)-->, data exploration, and finally in communicating the results.
<!-- role of making maps - from checking data, through the data exploration, to presenting the results  -->
<!-- maps in the computer age -->
Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.

All of the packages listed in Section \@ref(spatrep) have build-in methods for plotting spatial data using the `plot()` function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library.<!--this sentence can be easily removed-->
R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data.

### 2020-07-20 paper/paper.tex
adds a bunch of new ideas
adds creating maps section

@@

>>>>

\hypertarget{spatrep}{%
\subsection{Spatial data representations}\label{spatrep}}

Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
Each geometry is connected to a row in an attribute table.
A few hundred of different spatial data file formats were developed to store data in these formats.

While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????).
Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.

The most prominent package allowing for raster data representation is \emph{raster}.
A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed, however, it could take several years for this package to be adopted by users.
Alternatively, \emph{stars} package can be used to read and process raster data.
It is interconnected with the \emph{sf} package, and beside raster support, it provides tools for spatial data cubes.

The \emph{sp} package was a standard for vector data representation for more than 10 years.
As of 2020, more than 500 R packages directly depends or imports \emph{sp}.
However, it is now in a maintenance mode only with a recommended successor - the \emph{sf} package.

Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
They also have dedicated methods or functions to convert between representations, for example between \emph{sp} and \emph{sf} objects or \emph{raster} and \emph{stars} objects.
This can be useful when a given function expects different data representation than currently existing.

\hypertarget{spatial-data-download}{%
\subsection{Spatial data download}\label{spatial-data-download}}

@@

Creating maps is essential when working with spatial data.
It has a role in checking the quality of data, data exploration, and finally in communicating the results.
Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.

All of the packages listed in Section \ref{spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library.
R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data.
